---
title: "Modeling and prediction for movies"
output: 
  html_document: 
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

## Setup

### Load packages

```{r load-packages, message = FALSE, warning=FALSE}
library(ggplot2) # data visualization 
library(dplyr)   # data manipulation 
library(gridExtra) # multiple plots in one graph

# apply general knitr options
knitr::opts_chunk$set(comment=NA, fig.align='center')
```

* * *

## Part 1: Data
The movies dataset is comprised of 651 randomly sampled movies produced and release before 2016.  The dataset includes information from [Rotten Tomatoes](https://www.rottentomatoes.com/) and [IMDB](http://www.imdb.com/). Since the movies dataset is randomly sampled, statistical conclusion drawn from this dataset can be generalize to all movies produced and released before 2016.  However, movie ratings in the dataset only come from users of Rotten Tomatoes and users of IMDb.  Therefore, this dataset cannot reflect the opinions of all audiences and critics towards movies.  

The dataset has the character of an observational study.  With this dataset, we can only establish association but no causation.

The result is a data frame where each row contains a different movie and columns represent variables about the movies.  The dataset has 32 variables, among which some are only for information purposes and do not make any sense to include in a statistical analysis.  For example, 'actor1' through 'actor5', 'imdb_url' (link to IMDB page) and 'rt_url' (link to Rotten Tomatoes page). 

### Load data
```{r load-data}
# load the data set
load("movies.Rdata")
```

* * *

## Part 2: Research question
Is there an association between audience score and other variables including type of movie, runtime, imdb rating, critics scores, Oscar awards obtained (picture, actor, actress, director)? Answering this question helps to provide insight on what attributes makes a movie popular among audiences.  

* * *

## Part 3: Exploratory data analysis
```{r rating-dist, warning=FALSE}
summary(movies$audience_score)
ggplot(data=movies,aes(x=audience_score))+geom_histogram(binwidth=5)+xlab('Audience score') + ggtitle('Histogram of audience score')
```

Figure 1.  Distribution of audience score.  
The lowest score is 11.00 and the highest score is 97.00.  The mean of score is 62.36 and the median of score is 65.00.   

```{r categorical1-EDA, warning=FALSE,fig.width = 8}
f2a<-ggplot(movies,aes(x=genre))+geom_bar()+xlab('')+ggtitle('(a) Type of movie')+scale_x_discrete(labels = abbreviate) 
f2b<-ggplot(movies, aes(x = genre, y = audience_score)) + geom_boxplot() + xlab('')+ylab('Audience score') + ggtitle('(b) Audience score for each type of movie') + scale_x_discrete(labels = abbreviate)
grid.arrange(f2a,f2b,ncol=1)
```

Figure 2 (a) Barplot of movie types. (b) Boxplot of audience score for different movie types. Note that abbreviations of movie types are used in Figure 2(Ac&A: Action&Adventure; Anmt: Animation; AH&I: Art House&International; Cmdy: Comedy; Dcmn: Documentary; Hrrr: Horror; M&PA: Musical&Performing Arts; My&S: Mystery & Suspense; Othr: Other; SF&F: Science Friction & Fantasy).  

Figure 2a shows that the number of each type of movie varies a lot from each other.  The number of Drama movie is much higher than any other types of movie.  Figure 2b shows that the audience score of each type of movie varies from each other, indicating the possible relationship between movie type and movie popularity.  

```{r categorical2-EAD, warning=FALSE,fig.width = 8,fig.height=8}
f3a<-ggplot(movies, aes(x = best_pic_win, y = audience_score)) + geom_boxplot() + 
  ggtitle('(a) Whether or not the movie won \n a best picture Oscar (yes/no)') + 
  ylab('Audience score') + xlab("")
f3b<-ggplot(movies, aes(x = best_actor_win, y = audience_score)) + geom_boxplot() + 
  ggtitle('(b) Whether or not one of the  main actors \n in the movie ever won an Oscar (yes/no)') + 
  ylab('Audience score') + xlab("")
f3c<-ggplot(movies, aes(x = best_actress_win, y = audience_score)) + geom_boxplot() + 
  ggtitle('(c)Whether or not one of the main actresses \n in the movie ever won an Oscar (yes/no)') + 
  ylab('Audience score') + xlab("")
f3d<-ggplot(movies, aes(x = best_dir_win, y = audience_score)) + geom_boxplot() + 
  ggtitle('(d) Whether or not the director of\n the movie ever won an Oscar (yes/no)') + 
  ylab('Audience score') + xlab("")
grid.arrange(f3a,f3b,f3c,f3d,ncol=2)
```

Figure 3. (a) Boxplot of audience score for whether or not the movie won a best picture Oscar. (b) Boxplot of audience score for whether or not the main actors won an Oscar. (c) Boxplot of audience score for whether or not the main actresses won an Oscar. (d) Boxplot of audience score for whether or not the director won an Osacar.  

Figure 3 shows that the median audience score for movies that won the best-picture Oscar or for movies with Oscar awarded actresses/director is higher than the median audience score of movies without those. What's more, the difference between the median audience score of movies winning best-picture Oscar award and the median audience score of the ones without (Fig. 3a) is much larger than the difference in Fig.3c and Fig.3d, indicating the picture quality may play an important role in movie popularity.  

```{r categorical3-EAD, warning=FALSE,fig.width = 10}
f4a<-ggplot(data=movies,aes(x=critics_score,y=audience_score))+geom_jitter(alpha=0.5)+ggtitle("(a) Audience score v.s critics score") + geom_smooth(method = "lm") + xlab("Critics score") + ylab('Audience score')
f4b<-ggplot(data=movies,aes(x=imdb_rating,y=audience_score))+geom_jitter(alpha=0.5)+ggtitle("(b) Audience score v.s imdb rating") + geom_smooth(method = "lm") + xlab("imdb rating") + ylab('Audience score')
grid.arrange(f4a,f4b,ncol=2)
```

Figure 4. (a) Audience score vs. critics score. (b) Audience score vs. IMDB rating. Black dots are data points. The blue line is the model. The shaded gray area around the line tells us the variability we might expect in our predictions.  

Figure 4 shows that audience scores are potisively, linearly and strongly related to critics score/imdb rating, indicating the association between audience score (movie popularity) and critics scores/imdb rating.  

```{r}
movies %>% summarise(cor(imdb_rating, critics_score))
f5<-ggplot(data=movies,aes(x=imdb_rating,y=audience_score))+geom_jitter(alpha=0.5)+ggtitle("(a) Critics score vs. imdb rating") + geom_smooth(method = "lm") + xlab("imdb rating") + ylab('Critics score')
f5
```
Figure 5. Critics score vs. IMDB rating.

Figure 5 shows a strong linear correlation between critics score and imdb rating.  Therefore, those two variables are dependent on each other. To avoid collinearity problem, either critics score or imdb rating should be eliminated in model.  Figure 4b shows that the relationship between audience score and imdb rating is not as linear as the relationship between audience score and critics score.  Actually, some nonlinear relationship occurs between audience score and imdb rating.  Therefore, critics score rather than imdb rating will be used as the variable to model audience score.  

* * *

## Part 4: Modeling
The full initial model uses following variables to predict 'audience_score': 'genre', 'runtime', 'critics_score','best_pic_win','best_actor_win','best_actress_win','best_dir_win'. 'imdb_rating' is excluded due to its strong linear relationship with 'critics_score' (Fig. 5).  Also,'director','actor1' through 'actor5', 'imdb_url' and 'rt_url' are excluded since they are not meaningful for the determination of whether a movie will be popular or not. Although other variables such as 'studio' may also be meaningful, they are excluded in the linear model because they are beyond the interest of this study.  Here, P-value based backward elimination is used for model selection. Reasons why choosing this model selection method are: firstly it allows us to see which variables are statistically significant predictors of the response; secondly it requires fitting less models. 

Firstly, fitting data with the full model.  
```{r}
movies<-filter(movies,!is.na(runtime) & !is.na(critics_score) & !is.na(genre) 
             & !is.na(best_pic_win) & !is.na(best_actor_win) & !is.na(best_actress_win)
             & !is.na(best_dir_win))

m1<-lm(audience_score ~ genre + runtime + critics_score + best_pic_win + best_actor_win + best_actress_win + best_dir_win, data=movies)
summary(m1)
```

Variable 'best_dir_winyes' has the largest P-value (0.86 > 0.05), so we drop it, refit the model. Note that the categorical variable 'genre; has multiple levels. Although some of the levels yield much higher p-value than 'best_dir_winyes' does, we cannot drop those levels and keep other levels. Instead we keep all levels of "genre" since p-values of some of the levels are smaller than 0.05.   

```{r}
m2<-lm(audience_score ~ genre + runtime + critics_score + best_pic_win + best_actor_win + best_actress_win, data=movies)
summary(m2)
```

Variable 'best_actor_winyes' has the largest P-value (0.487 > 0.05), so we drop it, refit the model. Again, variable 'genre' is kept.    

```{r}
m3<-lm(audience_score ~ genre + runtime + critics_score + best_pic_win + best_actress_win, data=movies)
summary(m3)
```

Variable 'best_actress_win' has the largest P-value (0.349 > 0.05), so we drop it, refit the model. Again, variable 'genre' is kept.  

```{r}
m4<-lm(audience_score ~ genre + runtime + critics_score + best_pic_win, data=movies)
summary(m4)
```

Variable 'best_pic_win' has the largest P-value (0.324 > 0.05), so we drop it, refit the model. Again, variable 'genre' is kept.

```{r}
m5<-lm(audience_score ~ genre + runtime + critics_score, data = movies)
summary(m5)
```

Finally, the largest p-value is less than 0.05, then no variable should be dropped. Therefore, all predictors in the current model is our best-fitting model.  R-squared equals to 0.532, indicating about 53 precentage of variability in audience score is explained by the model.  P-value of the model is 2e-16, so the model as a whole is significant.  

Multiple regression methods depend on the following assumptions:  
1.Linear relationship between numerical explanatory variables and response variable.  
2.The residuals are nearly normal.  
3.The variability of the residuals is nearly constant.  
4.The residuals are independent.   
Here, diagnostic plots are used to check each of these assumptions.    

```{r,linear-relationship}
f6a<-ggplot(data=m5,aes(x=movies$runtime,y=.resid)) + geom_point() + ggtitle("(a) Residuals vs. runtime") + xlab("runtime (minute)") + ylab("Residuals") + geom_hline(yintercept = 0, linetype="dashed")
f6b<-ggplot(data=m5,aes(x=movies$critics_score,y=.resid)) + geom_point() + ggtitle("(b) Residuals vs. critics_score") + xlab("critics_score") + ylab("Residuals") + geom_hline(yintercept = 0, linetype="dashed")
grid.arrange(f6a,f6b,ncol=1)

```
Figure 6. (a) Residuals vs. runtime. (b) Residuals vs. critics_score.  
Figure 6 shows that residuals are randomly scattered about zero, indicating that the multiple regression model meets the first condition listed above.  

```{r, warning=FALSE}
ggplot(data = m5, aes(x=.resid)) + geom_histogram() + xlab("Residuals") + ggtitle("(a) Histogram of residuals")
qqnorm(m5$residuals,main = "(b) Normal probability plot of residuals") 
qqline(m5$residuals,col="red")
```
Figure 7. (a) Histogram of residuals. (b) Normal probability plot of residuals.  
Figure 7 shows that the residuals are normally distributed.  So the multiple regression model meets the second conition listed above.  

```{r,constant-residuals,warning=FALSE}
fig8a<-ggplot(data = m5, aes(x=.fitted,y=.resid)) + geom_point() + ggtitle("(a) Residuals vs. fitted") + xlab("Predicted values") + ylab("Residuals") + geom_hline(yintercept = 0, linetype="dashed")
fig8b<-ggplot(data = m5, aes(x=.fitted,y=abs(.resid))) + geom_point() + ggtitle("(b) Absolute value of residuals vs. fitted") + xlab("Predicted values") + ylab("Absolute value of residuals") + geom_hline(yintercept = 0, linetype="dashed")
grid.arrange(fig8a,fig8b,ncol=1)
```
Figure 8. (a) Residuals vs. predicted values. (b) Absolute value of residuals  

Figure 8a shows that residuals are randomly scattered in a band with a constant (40) with around zero. We can see a slight fan-shape in Figure 8a. Figure 8b is simply the Figure 8a folded in half.  Again, we can see the residuals are randomly scattered above zero with a bandwidth of about 40. The absolute value of residuals slightly decreases as predicted value increase. But the trend is not obvious.   We can conclude that the variability of the residual stays nearly constant as the predicted values change, so the constant variability condition is fairly satisfied. 

Independent residuals means independent observations.   We do not have any information about the order of data collection, so we do not have another diagnostic approch to check the independence condition.  However, since the movie data are randomly sampled and the sample size is less than 10% of the movie population, we do not have any reason to assume that the residuals are not independent.  Therefore, the residual independent condition is satisfiled.  

Therefore, all four conditions are met, indicating the credibility in the best-fit model.
The best-fit model shows that all else held constant, for each 1 increase in critics_score the model predicts the audience_score to be higher on average by 0.45; all else held constant, for each 1 minute increase in runtime the model predicts the audience_score to be higher on average by 0.07.  All else held constant, the order of the 11 levels of movie genre from lowest predicted audience score to highest predicted audience score is: Horror, Science Fiction & Fantasy, Mystery & Suspense, Comedy, Action and Adventure, Drama, Other, Art House & International, Animation, Documentary, Musical & Performing Arcts.  

* * *

## Part 5: Prediction
The best-fit model is then used to predict audience_score of the movie called "Me Before You" that released in 2016. The data for this movie come from [IMDB](http://www.imdb.com/) and [Rotten Tomatoes](https://www.rottentomatoes.com).  Runtime of this movie is 110 minute, its genre is Drama, critics_score is 59 and the audience_score is 77.
```{r, prediction, warning=FALSE}
newmovie<-data.frame(genre = "Drama", runtime=110, critics_score=59)
predict(m5,newmovie,interval = "prediction",level = 0.95)
```
The model predicts, with 95% confidence, that movie "Me before You" is expected to have an audience score between 63.84 and 91.31. The confidence interval includes the actual audience score of this movie (77). 

* * *

## Part 6: Conclusion
The metrics of movie popularity is complex. Here, I simpliy use audience score as a proxy for the popularity of a movie.  Using multiple linear regression model and a data set collected from [IMDB](http://www.imdb.com/) and [Rotten Tomatoes](https://www.rottentomatoes.com), we find that genre, runtime and critics score are statistically significant predictors of audience score.  While whether the movie won a best-picture Oscar or the director/actors/actresses won an Oscar are not significant predictors of audience socre.  However, this study only considers a subset of varibales for the initial modeling.  Variables such as title, mpaa_rating, studio etc. would also be significant predictors of audience score. Therefore, future study could consider other variables in the model.  



